---
title: "CamTrap"
author: "Carlos Tejeda Ocampo A011170840"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Analysis of camera trap metadata from Ocote Wildlife Reserve

```{r libraries}
library(camtrapR)
library(overlap)

```

## Verify exiftool executable

```{r verifyexiftool}
Sys.getenv("PATH")
Sys.which("exiftool")

```

## Load and view came data 

```{r loaddata}
cam_oco <- read.csv ("cam_oco.csv", header = TRUE) 

cam_oco #ver tabla archivo
names(cam_oco) #nombre de las variables del archivo

```

## Load photo metadata and add copyright tag

```{r loadmeta}
fotos_totales_uma_oco <- file.path("UMA_OCOTONES_CHIS")

proyecto_institucion <- "Proyecto_PN-2016/4106"

addCopyrightTag(inDir = fotos_totales_uma_oco, 
                copyrightTag = proyecto_institucion, 
                askFirst     = FALSE,
                keepJPG_original = F)

exifTagNames(fotos_totales_uma_oco, returnMetadata = TRUE)

```

## Extaract metadata

Extraction of metadata of each image corresponding to each species, associated to each site (camera-trap), in the first instance the general database must be taken out, to determine the total number of photographs

```{r maketable}
oco_tabla_general <- recordTable(inDir = fotos_totales_uma_oco,
                                              IDfrom = "metadata",
                                              metadataSpeciesTag = "Especies",
                                              timeZone = "America/Mexico_City",
                                              writecsv = FALSE)

head(oco_tabla_general)

write.csv(oco_tabla_general, "oco_tabla_general.csv")

```

The data table is generated with an independence of 12 h (it was the agreement that had to consider the data as independent), since the package works with minutes, it will be necessary to convert the hours to minutes. 12 h = 720 m.

```{r maketable2}
oco_tabla_de_independencias <- recordTable(inDir = fotos_totales_uma_oco,
                                                     IDfrom = "metadata",
                                                     metadataSpeciesTag = "Especies",
                                                     minDeltaTime = 720,
                                                     deltaTimeComparedTo = "lastRecord",
                                                     timeZone = "America/Mexico_City")

head(oco_tabla_de_independencias)


write.csv(oco_tabla_de_independencias, "oco_tabla de independencias.csv")


```

## Generate biodiversisty map

```{r maketable3}
cam_oco <- read.csv ("cam_oco.csv", header = TRUE) 

mapa_de_riqueza_de_especies_oco <- detectionMaps(CTtable     = cam_oco,
                                       recordTable  = oco_tabla_de_independencias,
                                       Xcol         = "utm_x",
                                       Ycol         = "utm_y",
                                       stationCol   = "Station",
                                       speciesCol   = "Species",
                                       printLabels  = TRUE,
                                       richnessPlot = TRUE,    
                                       speciesPlots = FALSE,
                                       addLegend    = TRUE)


```

## Generate camera activity report


```{r camreport}
actividad_de_camaras_oco <- cameraOperation(CTtable = cam_oco,
                                           stationCol = "Station",
                                           setupCol = "Setup_date",
                                           retrievalCol = "Retrieval_date",
                                           hasProblems = T,
                                           dateFormat = "%d/%m/%Y")

head(actividad_de_camaras_oco)
write.csv(actividad_de_camaras_oco, "actividad_de_camaras_oco.csv")

```


