---
title: "Music Information Retrieval"
author: "Carlos Tejeda Ocampo A011170840"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Music Information Retrieval

```{r libraries, warning=FALSE, message=FALSE}
library(tidyverse)
library(chorrrds)

library(chorddiag)

```

## Get chords from songs

Getting the song and select a random sample. Then selecting the chords of the songs.

```{r chords, echo=FALSE}

songs <- "the-beatles" %>% 
  chorrrds::get_songs() %>% 
  dplyr::sample_n(5)        

chords <- songs %>% 
  dplyr::pull(url) %>%                     
  purrr::map(chorrrds::get_chords) %>%     
                                           
  purrr::map_dfr(dplyr::mutate_if, is.factor, as.character)   %>% 
  chorrrds::clean(message = FALSE)        
                                            
                                           

chords %>% slice(1:10)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

## Calculate the most commong chords in each song


```{r modechord, echo=FALSE}

chords %>% 
  dplyr::group_by(song) %>% 
  dplyr::count(chord) %>%
  dplyr::top_n(n, n = 3) %>%
  dplyr::mutate(prop = scales::percent(n/sum(n)))
```

## Calculate the chord bigrams
Generate pairs of chords that happened in sequence, by song, and analyze their frequencies

```{r biagram, echo=FALSE}

chords %>%
  split(.$song) %>% 
  purrr:::map(chorrrds::chords_ngram, n = 2) %>% 
  dplyr::bind_rows() %>% 
  dplyr::group_by(song) %>% 
  dplyr::count(chords_ngram) %>% 
  dplyr::top_n(n, n = 2)
```

## Plot chord diagram

```{r diagram, echo=FALSE}

comp <- chords %>% 
  dplyr::mutate(seq = lead(chord)) %>% 
  dplyr::filter(chord != seq) %>% 
  dplyr::group_by(chord, seq) %>%  
  dplyr::summarise(n = n())

mat <- tidyr::spread(comp, key = chord, value = n, fill = 0)  
mm <- as.matrix(mat[, -1]) 

chorddiag::chorddiag(mm, showTicks = FALSE,
                     palette = "Blues")
```
